\documentclass[12pt]{article}

\input{packages.tex}
\input{style.tex}

\begin{document}

\title{Special Matrices and\\ Additional Properties of Matrices}
\author{Brian D.\ Fitzpatrick}
\date{\cite[\S1.4]{peterson}}

\maketitle

% \tableofcontents

\begin{sagesilent}
  latex.matrix_delimiters(left='[', right=']')
\end{sagesilent}

\begin{definition}
  A square matrix $A$ is \emph{diagonal} if $a_{ij}=0$ for $i\neq j$.
\end{definition}

\begin{sagesilent}
  A = matrix.diagonal([3,-1,1/2,0])
  B = matrix.diagonal([3,1])
\end{sagesilent}

\begin{ex}
  The matrices
  \begin{align*}
    \sage{A} && \sage{B} && I_n && 0_{n\times n}
  \end{align*}
  are diagonal.
\end{ex}

\begin{note}
  We write $\diag(\lambda_1,\lambda_2,\dotsc,\lambda_n)
  =
    \begin{bmatrix}
      \lambda_1 &           &        &           \\
      & \lambda_2 &        &           \\
      &           & \ddots &           \\
      &           &        & \lambda_n 
    \end{bmatrix}
  $
\end{note}

\newpage

\begin{thm}[Theorem 1.11 in \cite{peterson}]
  Let $A = \diag(a_1,\dotsc,a_n)$ and $B = \diag(b_1,\dotsc,b_n)$.  Then
  \begin{enumerate}
  \item $A+B=\diag(a_1+b_1,\dotsc,a_n+b_n)$
  \item $AB=\diag(a_1b_2,\dotsc,a_nb_n)$
  \item $A\in\GL_n(\RR)$ if and only if each $a_i\neq 0$. Furthermore, in this
    case, $A^{-1}=\diag(\nicefrac{1}{a_1},\dotsc,\nicefrac{1}{a_n})$
  \end{enumerate}
\end{thm}

\begin{definition}
  A square matrix $A$ is
  \begin{description}
  \item[\emph{upper triangular}] if $a_{ij}=0$ for $i>j$
  \item[\emph{lower triangular}] if $a_{ij}=0$ for $j>i$
  \end{description}
\end{definition}


\begin{sagesilent}
  A = matrix([[1,4,2],[0,0,4],[0,0,1]])
  B = matrix([[0,0,0],[2,8,0],[4,9,7]])
\end{sagesilent}
\begin{ex}
  We have
  \begin{align*}
    \underbrace{\sage{A}}_{\textnormal{upper triangular}} &&
    \underbrace{\sage{B}}_{\textnormal{lower triangular}}
  \end{align*}
\end{ex}

\begin{thm}[Theorem 1.12 in \cite{peterson}]
  Suppose $A$ and $B$ are upper triangular. Then
  \begin{enumerate}
  \item $A+B$ is upper triangular.
  \item $AB$ is upper triangular.
  \item $A$ is invertible if and only if each diagonal entry is nonzero.
  \end{enumerate}
  The corresponding theorem holds for lower triangular matrices.
\end{thm}

\newpage
\begin{definition}
  The \emph{transpose} of a $m\times n$ matrix $A$ is the $n\times m$ matrix
  $A^\top$ whose entries are $[A^\top]_{ij}=a_{ji}$.
\end{definition}

\begin{sagesilent}
  A = matrix([[1,2,3],[4,5,6]])
\end{sagesilent}
\begin{ex}
  $A=\sage{A}\implies A^\top=\sage{A.transpose()}$
\end{ex}

\begin{thm}[Theorem 1.13 in \cite{peterson}]\leavevmode
  \begin{enumerate}
  \item $(A^\top)^\top=A$
  \item $(A+B)^\top=A^\top+B^\top$
  \item $(\lambda A)^\top=\lambda A^\top$
  \item $(AB)^\top=B^\top A^\top$
  \item $(A^\top)^{-1}=(A^{-1})^\top$
  \end{enumerate}
\end{thm}
\begin{proof}
  To prove that $(AB)^\top=B^\top A^\top$, note that
  \begin{align*}
    [(AB)^\top]_{ij}
    &= [AB]_{ji} \\
    &= \sum_{k=1}^n [A]_{jk}[B]_{ki} \\
    &= \sum_{k=1}^n [A^\top]_{kj}[B^\top]_{ik} \\
    &= \sum_{k=1}^n [B^\top]_{ik}[A^\top]_{kj} \\
    &= [B^\top A^\top]_{ij}
  \end{align*}
  Hence $(AB)^\top=B^\top A^\top$.
\end{proof}


\newpage
\begin{definition}
  A matrix $A$ is \emph{symmetric} if $A^\top=A$. 
\end{definition}

\begin{idea}
  $\textnormal{rows of }A=\textnormal{columns of }A^\top$
\end{idea}

\begin{definition}
  We denote the collection of all $n\times n$ matrices by $\Sym_n(\RR)$.
\end{definition}

\begin{sagesilent}
  A = matrix([[1,2,3],[2,4,5],[3,5,6]])
  B = matrix([[1,2,3],[4,5,6],[7,8,9]])
\end{sagesilent}
\begin{ex}
  We have
  \begin{align*}
    \underbrace{\sage{A}}_{\textnormal{symmetric}} &&
    \underbrace{\sage{B}}_{\textnormal{not symmetric}}
  \end{align*}
\end{ex}


\begin{thm}[Theorem 1.14 in \cite{peterson}]\leavevmode
  \begin{enumerate}
  \item If $A$ and $B$ are symmetric, then $A+B$ is symmetric.
  \item If $A$ is symmetric, then $\lambda A$ is symmetric.
  \item $A^\top A$ and $A A^\top$ is symmetric.
  \item If $A$ is invertible and symmetric, then $A^{-1}$ is symmetric.
  \end{enumerate}
\end{thm}



\bibliography{mybib.bib}{}
\bibliographystyle{plain}

\end{document}
